% Many thanks to my advisor, <Your Advisor>, who took a chance on me four long years ago.  
% Through many pages of red ink and puzzling looks I have cut my teeth. 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% -- lead-in --
As I complete this dissertation, I stand on the shoulders of many people who have inspired and supported me over the course of my Ph.D.
First and foremost, I thank Alexa Lalejini, who (by request) gets their very own paragraph, and without whom, this dissertation would not have been possible. 
Thank you for moving to Michigan with me and making sure that I eat and sleep. 
Thank you for putting up with an apartment where the brick walls are absurdly cold in the winter and leak during rain storms because I thought it was an ``interesting'' place to live.
Thank you for helping me to take breaks by whisking me away on adventures. % that you plan.
Thank you for making sure that I eat and sleep.
Most of all, thank you for keeping me company. 

% -- friends and family --
Next, I thank my friends and family for their support over the years. 
I thank my parents, David and Penny Lalejini, for fostering my academic interests. % and supporting me over the years.
I especially want to acknowledge my dad who I could always count on for help with math homework or for someone to babble to about research projects.
I thank Brandon Odom for always being there for Alexa and me.
I also thank Josh Nahum for being my dependable exercise partner for the entire duration of my Ph.D. and for all of his friendship, kindness, and generosity. 

% -- academic mentors --
I have also had wonderful mentors, without whom I could not have bumbled my way through graduate school. % and meandered
As an undergraduate, I would not have been aware of graduate school as a realistic path for myself if not for Dr. Cindy Bethel at Mississippi State University who trusted me with research projects, funded trips to conferences, let me use her lab space after hours as a safe space to study for exams and do homework, and continues to be a font of advice and encouragement.

It has been a great privilege to have had Dr. Charles Ofria as my Ph.D. advisor and mentor.
Charles' immense excitement for research, teaching, and mentoring is infectious. 
Charles is one of my most valued role models in science, academia, and life.
I am also deeply appreciative of the supportive culture that Charles actively cultivates in the Digital Evolution Lab.
I want to specifically thank Dr. Anya Vostinar and Dr. Emily Dolson,
from each of whom I have learned so much about being a good researcher, academic, and person.
I also want to thank my committee---Dr. Christoph Adami, Dr. Wolfgang Banzhaf, and Dr. Richard Lenski---for their guidance and insightful feedback on my work. 

Finally, I want to acknowledge all of the members of the Digital Evolution Lab, past and present.
Our constructive discussions, collaborations, and social gatherings have enriched my work.
I especially want to thank
Dr. Mike Wiser,
Dr. Anya Vostinar,
Dr. Rosangela Canino-Koning,
Dr. Emily Dolson,
Anselmo Pontes,
Matthew Andres Moreno,
Jose Guadalupe Hernandez,
Austin J. Ferguson,
Acacia Ackles,
Kate Skocelas,
and Clifford Bohm.

% Bumbling
% - Mentors (bumble)
%   - Cindy Bethel
%   - Charles Ofria
%   - Dr. Anya Vostinar
%   - Dr. Emily Dolson - so much of learn by example, model my work on theirs
%   - Finally, committee---chris, wolfgang, and rich---...

% -- lab --
% Finally, 
% - (Ofria) Lab (+friends) + lab adjacent
%   - All members of the devolab, past and present
%   - Mike Wiser
%   - Rosangela Canino-Koning
%   - Anselmo Pontes
%   - Anya Vostinar
%   - Emily Dolson
%   - Matthew Andres Moreno
%   - Jose Guadalupe Hernandez
%   - Austin J. Ferguson
%   - Kate Skocelas
%   - Clifford Bohm
%   - constructive criticism, feedback, role modeling 

%%%%%%%%%%%%%%%%%%%%%%%%%%% NOTES %%%%%%%%%%%%%%%%%%%
% First person in family to navigate graduate school. Bumbled my way through. Would not even known how to bumble without my academic mentors along the way. I still don't know how to bumble the rest of the way, but I know I have a wealth of mentors to help.

% for all of the times that I phoned home for help with math homework or
% I especially want to thank my dad for all of the times that I phoned home for help with math homework or for an attentive ear to babble about research projects.

% Emily Dolson
% - My first experience writing a web-based visualization from scratch. Shout out to Emily Dolson for showing me how awesome web visualizations can be, for introducing me to D3js, and for letting me to pepper you with questions.

% Anya vostinar

% Explicitly call out co-authors for all thesis chapters.