\section{Discussion}

In this work, we used evolving populations of digital organisms to determine how adaptive phenotypic plasticity alters subsequent evolutionary dynamics and influences evolutionary outcomes in fluctuating environments.
First, we examined the evolutionary histories of plastic and non-plastic populations to test whether the evolution of adaptive plasticity promotes or constrains subsequent evolutionary change.
Next, we evaluated how adaptive plasticity influences fitness landscape exploration and exploitation by testing whether plastic populations are better able to evolve and then maintain novel traits.
Finally, we tested if the evolution of adaptive plasticity increases the potential for deleterious instructions to accumulate in evolving genomes.

Overall, our results indicate that adaptive plasticity can improve evolution's ability to maintain and refine novel traits, though with the tradeoff of reducing evolutionary exploration of the fitness landscape.
Additionally, we found no evidence that adaptive plasticity increased the potential for deleterious instructions to accumulate in genomes.
Instead, the genomes of non-plastic organisms that evolved in an identical fluctuating environment accumulated more deleterious instructions than that of adaptively plastic genomes. 
These dynamics appear to be driven by the stabilizing effect that adaptive plasticity had on population dynamics rather than plasticity's effect on genetic architecture or regulation.

\subsection{The speed of evolutionary change}

% -- Adaptively plastic populations underwent less evolutionary change than non-plastic populations --
Adaptively plastic populations experienced fewer selective sweeps and fewer total genetic changes relative to non-plastic populations evolving under the same environmental conditions (Figure \ref{chapter:consequences-of-plasticity:fig:evolutionary-dynamics-magnitude}).
Plastic populations adapted to the fluctuating environmental regime by evolving to sense environmental changes and regulate their metabolism (task performance) in response to such changes, which also stabilized these populations against fluctuations.
Indeed, across all three of our experiments, the evolutionary dynamics of plastic populations were more similar to that of populations evolving in a static environment than to that of non-plastic populations evolving in an identical fluctuating environment.

% -- Non-plastic populations repeatedly re-adapted to environmental change --
Adaptive phenotypes in ENV-A were maladaptive in ENV-B and vice versa.
As such, non-plastic generalists that performed all tasks or performed no tasks at all did not evolve in any of the fluctuating environmental regimes.
Selection against non-plastic generalists may be attributed to competition with phenotypic specialists that have a much larger fitness advantage for performing environment-specific tasks.
In non-plastic populations where plasticity was disallowed, we hypothesize that strong selection on task specialization after each environmental change drove the repeated fixation of beneficial mutations (that alter an organism's phenotypic profile).
This hypothesis is supported by the increased frequency of coalescence events in these populations (Figure \ref{chapter:consequences-of-plasticity:fig:evolutionary-dynamics-rate}) as well as increased rates of genetic and phenotypic changes observed along the lineages of non-plastic organisms. 

% -- architecture results --
Analysis of the evolved genetic architectures further supports our hypothesis that the non-plastic populations relied on mutations to continuously readapt to the fluctuating environment. 
This aligns with previous work, which has shown that, in the absence of plasticity, fluctuating environments steer populations toward genotypes that readily mutate to alternative phenotypes \citep{lalejini_evolutionary_2016, canino-koning_evolution_2016}.
Indeed, \citep{canino-koning_evolution_2016} also observed that genomes evolved in cyclic environments often contained vestigial fragments of genetic material adapted to prior environments, which we see in the non-plastic populations. 

% -- in context of previous digital evolution work --
This study is the first in-depth empirical investigation into how the \textit{de novo} evolution of adaptive plasticity shifts the course of subsequent evolution in a cyclic environment.
The evolutionary dynamics that we observed in non-plastic populations, however, are consistent with results from previous digital evolution studies. 
Consistent with our findings, \cite{dolson_interpreting_2020} showed that non-plastic populations that were evolved in cyclically changing environments exhibited higher phenotypic volatility and accumulated more mutations than that of populations evolved in static conditions.
%\cite{lalejini_evolutionary_2016} visually inspected the evolutionary histories of non-plastic organisms evolved in fluctuating environments, observing that mutations readily switched the set of traits expressed by offspring.
%\cite{canino-koning_evolution_2016} investigated how different types of changing environments shape the genetic architectures of evolved organisms, showing that cyclically-changing environments can steer populations toward genotypes that more readily mutate to alternative phenotypes.
%\cite{canino-koning_evolution_2016} also observed that genomes evolved in harsh cyclic environments often contained vestigial fragments of genetic material adapted to prior environments.
%  - @Austin - depending on what you include about architecture results, you'll also want to work in results from Canino-Koning (2019); going to talk about their novel tasks results in next subsection.


% -- In context of conventional evolutionary theory: evo response => f(selection, variation) --
Our results are also consistent with conventional evolutionary theory.
A trait's evolutionary response to selection depends on the strength of directional selection and on the amount of genetic variation for selection to act upon \citep{lande_measurement_1983,zimmer_evolution_2013}.
In our experiments, non-plastic populations repeatedly experienced strong directional selection to toggle which tasks were expressed after each environmental change.
As such, retrospective analyses of successful lineages revealed rapid evolutionary responses (that is, high rates of genetic and phenotypic changes).
Evolved adaptive plasticity shielded populations from strong directional selection when the environment changed by eliminating the need for a rapid evolutionary response to toggle task expression.
Indeed, both theoretical and empirical studies have shown that adaptive plasticity can constrain evolutionary change by weakening directional selection on evolving populations \citep{price_role_2003,paenke_influence_2007,ghalambor_non-adaptive_2015}. 

\subsection{The evolution and maintenance of novel tasks}

% -- Exploration + Exploitation --
In fluctuating environments, non-plastic populations explored a larger area of the fitness landscape than adaptively plastic populations, as measured by novel task discovery (Figure \ref{chapter:consequences-of-plasticity:fig:complex-traits-magnitude}b).
Despite lower overall novel task discovery in adaptively plastic populations, they better exploited the fitness landscape, retaining a greater number of novel tasks than non-plastic populations evolving under identical environmental conditions (Figure \ref{chapter:consequences-of-plasticity:fig:complex-traits-magnitude}a).
Evolution in non-plastic populations was dominated by numerous bouts of strong directional selection driven by repeated environmental change.
After each change, the performance of the six base tasks needed to be realigned to the environment. 
In our experiment, novel tasks were less important to survival than the fluctuating base tasks.
In non-plastic populations, mutations that improve an offspring's fitness after an environmental change are extremely beneficial, and as such beneficial mutations fix, they can carry with them co-occurring deleterious mutations that knock out novel tasks. 
Indeed, we found that mutations associated with novel task loss along representative lineages from non-plastic populations co-occurred with mutations that helped offspring adapt to environmental changes 97\% of the time.

% -- Changing environments promote evolutionary change --
Temporary environmental changes can improve fitness landscape exploration and exploitation in evolving populations of non-plastic digital organisms \citep{nahum_improved_2017}.
In our system, however, we found that \textit{repeated} fluctuations reduced the ability of non-plastic populations to maintain and exploit tasks; that said, we did find that repeated fluctuations may improve overall task discovery by increasing generational turnover. 
Consistent with our findings, \cite{canino-koning_fluctuating_2019} found that non-plastic populations of digital organisms evolving in a harsh cyclic environment maintained fewer novel traits than populations evolving in static environments.

% -- Plastic rescue, stabilizing effect of plasticity --
Our results suggest that adaptive phenotypic plasticity can improve the potential for populations to exploit novel resources by stabilizing them against stressful environmental changes.
The stability that we observe may also lend some support to the hypothesis that phenotypic plasticity can rescue populations from extinction under changing environmental conditions \citep{chevin_adaptation_2010}.

% -- relevance to genes as followers hypothesis --.
Our data do not necessarily provide evidence for or against the genes as followers hypothesis.
The genes as followers hypothesis focuses on contexts where plastic populations experience novel or abnormally stressful environmental change.
However, in our system, environmental changes were cyclic (not novel), and the magnitude of changes were consistent for the entirety of the experiment (so none were abnormally stressful).
Further, the introduction of novel tasks during the second phase of the experiment merely added additional static opportunities for fitness improvement and did not change the meaning of existing environmental cues. 

\subsection{The accumulation of deleterious instructions}

% -- Overview of results --
%   - More accumulation in non-plastic
%   - no evidence for cryptic variation housing poison
%   - plastic ~~ static
We found that non-plastic lineages that evolved in a fluctuating environment exhibited both larger totals and higher rates of deleterious instruction (\code{poison}) accumulation than that of adaptively plastic lineages (Figure \ref{chapter:consequences-of-plasticity:fig:deleterious-hitchhiking}).
We did not find evidence of \code{poison} instructions accumulating as cryptic variation in adaptively plastic lineages.
We hypothesize that deleterious genetic hitchhiking drove \code{poison} instruction accumulation along non-plastic lineages in changing environments.
In asexual populations without horizontal gene transfer, all co-occurring mutations are linked.
As such, deleterious mutations linked with a stronger beneficial mutation (\textit{i.e.}, a driver) can sometimes ``hitchhike'' to fixation \citep{smith_hitch-hiking_1974,van_den_bergh_experimental_2018,buskirk_hitchhiking_2017}.
Natural selection normally prevents deleterious mutations from reaching high frequencies, as such mutants would be outcompeted.
However, when a beneficial mutation sweeps to fixation in a clonal population, it carries along any linked genetic material, including other beneficial, neutral, or deleterious mutations  \cite{barton_genetic_2000,smith_hitch-hiking_1974}.

Across our experiments, the frequency of selective sweeps in non-plastic populations provided additional opportunities for genetic hitchhiking with each environmental change. 
Indeed, representative lineages from non-plastic populations in the cyclic environment exhibited higher mutation accumulation (Figure~\ref{chapter:consequences-of-plasticity:fig:evolutionary-dynamics-magnitude}b), novel trait loss (Figure~\ref{chapter:consequences-of-plasticity:fig:complex-traits-magnitude}c), and deleterious instruction accumulation (Figure~\ref{chapter:consequences-of-plasticity:fig:deleterious-hitchhiking}) than their plastic counterparts.
In aggregate, we found that many ($\sim$49\%; 956 / 1916) mutations that increased \code{poison} instruction execution in offspring co-occurred with mutations that provided an even stronger benefit by adapting the offspring to an environmental change.
This rate of co-occurrence is conservative because we did not analyze mutations that became linked in different generations.

% Phenotypic plasticity can be encoded in genomes in two ways: ...

% -- Elaboration on plastic result --
We found that adaptive phenotypic plasticity reduced \code{poison} instruction accumulation by reducing the rate of evolutionary change, which in turn reduced opportunities for \code{poison} instructions to fix.
We did not find compelling evidence of cryptic variation harboring \code{poison} instructions in adaptively plastic lineages.
We have two hypotheses for why we did not observe the accumulation of \code{poison} instructions in unexpressed plastic responses.
First, the period of time between environmental changes was too fast for variants carrying unexpressed \code{poison} instructions to reach high frequencies before the environment changed, after which such variants would have been outcompeted.
Indeed, we tuned the frequency of environmental fluctuations so that genes that needed to function appropriately in the off environment were able to remain in the population despite relaxed selection.
Second, the genetic mechanisms of plasticity that evolve in Avida are typically well-integrated and highly specific; that is, plastic genomes usually adjust their phenotypic expression by toggling a minimal number of key instructions. 
As such, there is little genomic space for variation to accumulate in preexisting (but unexpressed) regulated regions.

\subsection{Limitations and future directions}

% -- Adaptive vs non-adaptive plasticity --
Our conclusions are limited to \textit{adaptively} plastic populations.
We did not explore the effects of non-adaptive plasticity where environmental changes induce phenotypes that are further away from the local optimum (\textit{e.g.}, \citealt{leroi_temperature_1994}).
Non-adaptive plasticity can increase a population's extinction risk, especially if the misaligned plastic response is strongly tied to survival or the population is not sufficiently large \citep{gomulkiewicz_when_1995,chevin_adaptation_2010}.
If the population persists, however, non-adaptive plasticity has been shown to be capable of accelerating evolutionary change by increasing the strength of directional selection. \citep{ghalambor_non-adaptive_2015}.

% -- Environmental change --
Environmental cues in our experiments were reliable, and environmental changes were consistent over time. 
That is, sensory instructions perfectly differentiated between ENV-A and ENV-B, and environmental fluctuations never exposed populations to entirely new conditions.
Both the reliability of cues and the timescales of environment switching are known to influence evolutionary outcomes \citep{li_digital_2004,boyer_adaptation_2021}. % [cite - cue reliability].
For example, \cite{boyer_adaptation_2021} evolved populations of \textit{Saccharomyces cerevisiae} in an environment that fluctuated between two growth conditions, observing that both environmental predictability and switching rate influenced the rates of evolutionary responses as well as adaptive outcomes. 
In adaptively plastic populations, environmental switching rate can influence how plastic responses are maintained, including their genetic architecture as well as their likelihood of maintenance. 
Our work lays the groundwork for using digital evolution experiments to investigate the evolutionary consequences of phenotypic plasticity in a range of contexts, including different forms of plasticity (\textit{e.g.}, adaptive versus non-adaptive), more complex environments with more than two possible states, stochastic environmental changes, and different environment switching rates.

% - limitation: focus on lineages -
We focused our analyses on the lineages of organisms with the most abundant genotype in the final population.
These successful lineages represented the majority of the evolutionary histories of populations at the end of our experiment, as populations did not exhibit long-term coexistence of different clades.
Our analyses, therefore, gave us an accurate picture of what fixed in the population.
We did not, however, examine the lineages of extinct clades.
Future work will extend our analyses to include extinct lineages, giving us a more complete view of evolutionary history, which may allow us to better distinguish adaptively plastic populations from populations evolving in a static environment. 

% - Machinery -
As with any wet-lab experiment, our results are in the context of a particular model organism: ``Avidian'' self-replicating computer programs.
Digital organisms in Avida regulate responses to environmental cues using a combination of sensory instructions and conditional logic instructions (\code{if} statements).
The \code{if} instructions conditionally execute a single instruction depending on previous computations and the state of memory.  
As such, plastic genomes typically regulate a small number of key instructions that, when executed, change the expressed phenotype as opposed to large, sequences of co-regulated instructions \citep{consequences_of_plasticity_supplemental_material_2021}.
This bias may limit the accumulation of hidden genetic variation in Avida genomes. 
However, as there are many model biological organisms, there are many model digital organisms that have different regulatory mechanisms that should be used to test the generality of our results.

% [A broad sentence to wrap everything up?]
% e.g., A sentence about how we hope this work inspires more use of digital evolution as an experimental tool/expand experimental repertoire of evolutionary biologists studying phenotypic plasticity?
% e.g., As demonstrated here, Digital evolution studies allow us to directly manipulate the capacity for plastic responses to evolve and perfectly observe subsequent dynamics, enabling us to experimentally test hypotheses that were previously relegated to theoretical analyses.

